A Hybrid Ensemble Learning Approach to Star-Galaxy Classification

### 2015-07-10 paper/main.tex
the submitted version of the paper
update paper

\title{A Hybrid Ensemble Learning Approach to Star-Galaxy Classification}

@@

\label{firstpage}
\begin{abstract}
%Star-galaxy separation in currently ongoing and future
%large photometric surveys, such as DES and LSST,
%will detect a vast number of unresolved galaxies at faint magnitudes.
There exist a variety of star-galaxy classification techniques,
each with their own strengths and weaknesses.
In this paper, we present a novel meta-classification
framework that combines and fully exploits different techniques
to produce a more robust star-galaxy classification.
To demonstrate this hybrid, ensemble approach,
we combine a purely morphological classifier,
a supervised machine learning method based on random forest,
an unsupervised machine learning method based on self-organizing maps,
and a hierarchical Bayesian template fitting method.
%Any of these four methods could
%easily be replaced by a similar method
%or be extended to include other existing techniques.
Using data from the CFHTLenS survey,
we consider different scenarios:
when a high-quality training set is available with spectroscopic labels from
DEEP2, SDSS, VIPERS, and VVDS, and
when the demographics of sources in a low-quality training set
do not match the demographics of objects in the test data set.
We demonstrate that our Bayesian combination technique improves
the overall performance over any individual classification method
in these scenarios.
Thus, strategies that combine the predictions of different classifiers
may prove to be optimal in currently ongoing and forthcoming
photometric surveys,
such as the Dark Energy Survey and the Large Synoptic Survey Telescope.

@@

The problem of source classification is fundamental to astronomy
and goes as far back as \citet{messier1781catalogue}.
A variety of different strategies have been developed 
to tackle this long-standing problem,
%and yet it remains an unsolved challenge.
and yet there is no consensus on
the optimal star-galaxy classification strategy.
The most commonly used method to classify stars and galaxies
in large sky surveys is the morphological separation
~\citep{sebok1979optimal, kron1980photometry, valdes1982resolution,
yee1991faint, vasconcellos2011decision,
henrion2011bayesian}.
It relies on the assumption that
stars appear as point sources
while galaxies appear as resolved sources.
However,
currently ongoing and upcoming large photometric surveys,
such as the Dark Energy Survey
(DES\footnote{http://www.darkenergysurvey.org/})
and the Large Synoptic Survey Telescope
(LSST\footnote{http://www.lsst.org/lsst/}),
will detect a vast number of unresolved galaxies
at faint magnitudes.
Near a survey's limit, the photometric observations
cannot reliably separate stars from unresolved galaxies
by morphology alone without leading to
incompleteness and contamination in the star and galaxy samples.

@@

A self-organizing map~\citep{kohonen1990self, kohonen2001self}
is an unsupervised, artificial neural network algorithm
that is capable of projecting high-dimensional input data
onto a low-dimensional map
through a process of competitive learning.
In astronomical applications,
the high-dimensional input data can be
magnitudes, colors, or some other photometric attributes.
The output map is usually chosen to be two-dimensional
so that the resulting map can be used for visualizing
various properties of the input data.
The differences between a SOM and other neural network algorithms are
that a SOM is unsupervised,
there are no hidden layers and therefore no extra parameters,
and it produces a direct mapping
between the training set and the output network.
In fact, a SOM can be viewed as a non-linear generalization
of a principal component analysis (PCA) algorithm~\citep{yin2008self}.

@@

During the construction of a SOM,
each node on the two-dimensional map is represented by
weight vectors of the same dimension
as the number of attributes used to create the map itself.
In an iterative process,
each object in the input sample is individually used
to correct these weight vectors.
This correction is determined so that the specific neuron (or node),
which at a given moment best represents the input source,
is modified along with the weight vectors
of that node's neighboring neurons.
As a result, this sector within the map
becomes a better representation of the current input object.
This process is repeated for every object in the training data,
and the entire process is repeated for several iterations.
Eventually, the SOM converges to its final form where
the training data is separated into groups of similar features.

@@

One of the most important parameter in SOMc
is the topology of the two-dimensional SOM,
which can be rectangular, hexagonal, or spherical.
To classify stars and galaxies in the CFHTLenS data,
we use a spherical topology,
which is constructed by using
\textsc{HEALPIX}~\citep{gorski2005healpix}.
Furthermore, similar to TPC,
we use the OOB technique to make an unbiased estimation of errors.
For a complete description of the SOM implementation
and its application to the estimation of
photo-$z$ probability density functions (photo-$z$ PDFs),
we refer the reader to \cite{carrascokind2014somz}.

@@

We therefore adopt a binning strategy similar to 
\cite{carrascokind2014exhausting}.
In this binning strategy,
we allow different classifier combinations 
in different parts of parameter space
by creating two-dimensional SOM representations of
the full nine-dimensional magnitude-color space:
$u$, $g$, $r$, $i$, $z$, $u-g$, $g-r$, $r-i$, and $i-z$.
A SOM representation can be rectangular, hexagonal, or spherical;
here we choose a 10$\times$10 rectangular topology to facilitate 
visualization as shown in Figure~\ref{fig:som_colors}.
For all combination methods,
we use only the OOB (cross-validation) data contained in each cell
to compute the relative weights for the base classifiers.
The weights within individual cells are then applied to
the blind test data set to make the prediction.

@@

We have cross-matched reliable spectroscopic galaxies from
the Deep Extragalactic Evolutionary Probe Phase 2~
\citep[DEEP2;][]{davis2003science,newman2013deep2},
the Sloan Digital Sky Survey Data Release 10~\citep[SDSS-DR10]{Ahn2014},
the VIsible imaging Multi-Object Spectrograph (VIMOS)
Very Large Telescope (VLT) Deep Survey~
\citep[VVDS;][]{le2005vimos,garilli2008vimos}, and
the VIMOS Public Extragalactic Redshift
Survey~\citep[VIPERS;][]{garilli2014vimos},
In the end, we have 8,545 stars and 57,843 galaxies available
for the training and testing processes.
We randomly select 13,278 objects for the blind testing set,
and use the remainder for training and cross-validation.

Our goal here is not to obtain the best classifier
performance; for this we would have fine tuned individual base
classifiers and chosen sophisticated models best suited to the
particular properties of the CFHTLenS data.
For example, \cite{hildebrandt2012cfhtlens} suggest that
all objects with $i > 23$ in the CFHTLenS data set
may be classified as galaxies
without significant incompleteness and contamination in the galaxy sample.
Although this approach works because
the high Galactic latitude fields of the CFHTLS contain
relatively few stars,
it is very unlikely that such an approach
will meet the science requirements for the quality of
star-galaxy classification in lower-latitude, star-crowded fields.
Rather, our goal for the CFHTLenS data set
is to demonstrate the usefulness
of combining different classifiers even when
the base classifiers may be poor or trained on partial data.

@@

%% rich training data
\begin{table*}
  \caption{A summary of the classification performance metrics
           for the four individual methods
           and the four different classification combination methods
           as applied to the CFHTLenS data,
           with no cut applied to the training data set.
           The definition of the metrics is summarized in
           Table~\ref{table:metrics}.
           The bold entries highlight the best performance values
           within each column.}
  \centering
  \input{tables/metrics_all.tex}
  \label{table:metrics_all}
\end{table*}

@@

In Figure~\ref{fig:purity_mag}, 
we compare the star and galaxy purity values
for BMC, TPC, and morphological separation
as functions of $i$-band magnitude for the differential counts.
We use the kernel density estimation \cite[KDE;][]{silverman1986density}
with the Gaussian kernel to smooth the fluctuations in the distribution.
Although morphological separation shows a slightly better performance
in galaxy purity at bright magnitudes $i \la 20$,
BMC outperforms both TPC and morphological separation
at faint magnitudes $i \ga 21$.
As the top panel shows,
the number count distribution peaks at $i \sim 22$,
and BMC therefore outperforms both TPC and morphological separation
for the majority of objects.
It is also clear that BMC outperforms TPC over all magnitudes.
BMC can presumably accomplish this by combining information from
all base classifiers,
\eg giving more weight to the morphological separation method
at bright magnitudes.
The bottom panel shows that
the star purity of morphological separation drops to $p_s < 0.8$
at fainter magnitudes $i > 21$.
This is expected, as our crude morphological separation classifies
every object as a galaxy beyond $i > 21$, and
purity measures the number of true stars classified as stars.
It is again clear that BMC outperforms both TPC and morphological separation
in star purity values over all magnitudes.

In Figure~\ref{fig:purity_mag_integrated}, we show
the overall galaxy and star purity values as functions of magnitude
for the integrated counts.
Although morphological separation performs better than TPC
at bright magnitudes, its purity values decrease as 
the magnitudes become fainter, and
TPC eventually outperforms morphological separation by 1--2\% at
$i > 21$.
BMC clearly outperforms both TPC and morphological separation,
and it maintains the overall galaxy purity of 0.980
up to $i \sim 24.5$.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/purity_mag.pdf}
  \caption{Purity as a function of the $i$-band magnitude
           as estimated by the kernel density estimation (KDE) method
           for the differential counts.
           The top panel shows the histogram with a bin size of 0.1 mag
           and the KDE for objects in the test set.
           The second panel shows the fraction of stars estimated by KDE
           as a function of magnitude.
           The bottom two panels compare
           the galaxy and star purity values for BMC, TPC, and 
           morphological separation as functions of magnitude.
           Results for BMC, TPC, and morphological separation are in
           blue, green, and red, respectively.
           The $1 \sigma$ confidence bands are estimated by
           bootstrap sampling.}
  \label{fig:purity_mag}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/purity_mag_integrated.pdf}
  \caption{Purity as a function of the $i$-band magnitude
           for the integrated counts.
           The upper panel compares
           the galaxy purity values for BMC (blue solid line),
           TPC (green dashed line), and 
           morphological separation (red dashed line).
           The lower panel compares the star purity.
           The $1 \sigma$ error bars are computed following the method
           of \citet{paterno2004calculating} to avoid the unphysical
           errors of binomial or Poisson statistics.}
  \label{fig:purity_mag_integrated}
\end{figure}

@@

\begin{table*}
  \caption{A summary of the classification performance metrics
           for the four individual methods
           and the four different classification combination methods
           when the training data set consists of
           only the sources that are in CFHTLS W1 field,
           has spectroscopic labels available from VVDS,
           and has $i < 22$.
           The definition of the metrics is summarized in
           Table~\ref{table:metrics}.
           The bold entries highlight the best performance values
           within each column.}
  \centering
  \input{tables/metrics_cut.tex}
  \label{table:metrics_cut}
\end{table*}

We present in Table~\ref{table:metrics_cut}
the same six metrics for each method,
and highlight the best method for each metric.
Overall, the results obtained for the reduced data set are remarkable.
With a smaller training set, our training based methods, TPC and SOMc,
suffer a significant decrease in performance.
The performance of morphological separation and HB
is essentially unchanged from Table~\ref{table:metrics_all}
as they do not depend on the training data.
Without sufficient training data,
the advantage of combining the predictions of different classifiers
is more obvious.
Even WA, the simplest of combination techniques, outperforms
all individual classification techniques in four metrics,
AUC, $p_s$ at $c_s=0.7145$, $p_g$ at $c_g=0.9600$, and $p_s$ at $c_s=0.2500$.
%Interestingly, although TPC does not perform better
%than morphological separation,
%BoM still chooses TPC as the best model within each SOM cell
%and shows exactly the same performance metrics as TPC.
%Stacking and BMC techniques still outperform
%all individual classification techniques,
Although BoM always chooses TPC as the best model
when we have a high-quality training set,
it now chooses various methods in different bins
and outperforms all base classifiers.
While the performance of the stacking technique is only slightly worse
than that of BMC when we have a high-quality training set,
stacking now fails to outperform morphological separation.
BMC shows an impressive performance and 
outperforms all other classification techniques in all six metrics.
Overall, the improvements are small but still significant
since these metrics are averaged over the full test data.

%% ROC curve for vvds i < 20
%\begin{figure}
%  \centering
%  \includegraphics[width=\columnwidth]{figures/roc_cut.eps}
%  \caption{A magnified view of the ROC curve of
%           four star-galaxy individual classification approaches
%           (TPC, SOMc, HB, and morphology)
%           and the best-performing combination technique (BMC).
%           trained with the reduced data set.
%           We show only one of the four combination methods, BMC,
%           which has the best overall performance.}
%  \label{fig:roc_cut}
%\end{figure}

%In Figure~\ref{fig:roc_cut}, we plot the ROC curve
%of all four individual classification algorithm
%and our BMC technique for the poor training data case.
%We note that a binary classifier,
%such as our cut-based morphological separation,
%whose output is either zero or one,
%is a single point in the ROC space.
%We vary the half-light radius to create a smooth ROC curve
%for morphological separation.
%It is clear that our BMC technique dominates
%each individual classification techniques in the ROC space.

%% weights map for vvds i < 20

@@

In Figure~\ref{fig:weights_cut}, we again show the $10\times10$
two-dimensional weight map defined by the SOM.
%It seems that TPC performs well
%in most of the parameter space.
When the quality of training data is relatively poor,
the performance of training based algorithms will decrease,
while the performance of template fitting algorithms
or morphological separation methods
is independent of training data.
Thus, when the weight maps of Figure~\ref{fig:weights}
and Figure~\ref{fig:weights_cut} are visually compared,
it is clear that
the BMC algorithm now uses more information from
morphological separation and HB,
while it uses considerably less information from 
our training based algorithms, TPC and SOMc.
Not surprisingly, the morphological separation method
performs best at bright magnitudes,
and BMC assigns more weight to HB at fainter magnitudes.

We present the star and galaxy purity values as functions of
$i$-band magnitude for the differential counts
in Figure~\ref{fig:purity_mag_cut}.
The normalized density distribution as a function of magnitude
in the top panel
and the stellar distribution in the second panel
clearly show that the demographics of the training set and 
that of the test set are different.
Since the training set is cut at $i < 22$,
the density distribution falls off sharply around $i \sim 22$
and has a higher fraction of stars than the test set.
Compared to the purity values in Figure~\ref{fig:purity_mag},
TPC now suffers a significant decrease in star and galaxy purity.
However, the purity of BMC does not show such a significant drop
and decreases by only 2--5\%.
As suggested by the weight maps in Figure~\ref{fig:weights_cut},
BMC can accomplish this by shifting the relative weights assigned to
each base classifier in different SOM cells.
As the quality of training set worsens,
BMC assigns less weight to training based methods
and more weight to HB and morphological separation.

In Figure~\ref{fig:purity_mag_cut_integrated}, we show
the overall galaxy and star purity values as functions of magnitude
for the integrated counts.
Compared to Figure~\ref{fig:purity_mag_integrated},
the drop in the performance of TPC is clear.
However, even when some classifiers have been trained on
a significantly reduced training set,
BMC maintains a galaxy purity of 0.970 and a star purity of 1.0
up to $i \sim 24.5$, and it sill outperforms morphological separation
at fainter magnitudes $i \ga 21$.

@@

%\begin{figure*}
%  \centering
%  \includegraphics[width=\columnwidth]{figures/color_color.eps}
%  \caption{Color-color diagrams of sources with
%           $0.45<P\left(S\right)<0.55$ when the BMC technique
%           is applied to the poor training data.}
%  \label{fig:color_color}
%\end{figure*}
%
%To identify the regions in color-color space
%where our BMC technique struggles to classify stars and galaxies,
%we present the distribution over the color-color space for sources with
%$0.45 < P(S) < 0.55$ in Figure~\ref{fig:color_color}.
%Not surprisingly, these regions correspond to
%where there is a significant overlap between stars and galaxies
%in the color space.
%It may be possible to improve the performance of
%star-galaxy classification by obtaining more information
%through spectroscopic follow-up, manual study, or citizen science projects,
%or by developing a classification technique specifically designed
%for these regions of the color space and
%combining it with other techniques in a Bayesian framework.

@@

Similarly, we impose various spectroscopic redshift cuts 
on the training data in Figure~\ref{fig:perform_z_cut}. 
BMC begins to perform worse than morphological separation
when a conservative cut of $z_{\rmn{spec}} < 0.6$ is imposed.
However, it is again clear that BMC is able to
utilize information from HB and morphological separation to
mitigate the drop in the performance of TPC.

@@

The combination technique described in this paper
can be easily applied to other surveys.
Given the efficacy of our approach,
classifier combination strategies are likely the optimal approach
for currently ongoing and forthcoming photometric surveys.
We therefore plan to apply the combination technique described in this paper
to other surveys such as the DES.
Our approach can also be extended more broadly to
classify objects that are neither stars nor galaxies (\eg quasars).
Finally, future studies could explore the use of multi-epoch data,
which would be particularly useful for the next generation of
synoptic surveys.

@@

We thank Ignacio Sevilla for helpful and insightful conversations.
We acknowledge support from the 
National Science Foundation Grant No.\ AST-1313415.
RJB acknowledges support as an Associate
within the Center for Advanced Study at the University of Illinois.

>>>>

\title[A Hybrid Approach to Star-Galaxy Classification]{A Hybrid Ensemble Learning Approach to Star-Galaxy Classification}

@@

\label{firstpage}
\begin{abstract}
There exist a variety of star-galaxy classification techniques,
each with their own strengths and weaknesses.
In this paper, we present a novel meta-classification
framework that combines and fully exploits different techniques
to produce a more robust star-galaxy classification.
To demonstrate this hybrid, ensemble approach,
we combine a purely morphological classifier,
a supervised machine learning method based on random forest,
an unsupervised machine learning method based on self-organizing maps,
and a hierarchical Bayesian template fitting method.
Using data from the CFHTLenS survey,
we consider different scenarios:
when a high-quality training set is available with spectroscopic labels from
DEEP2, SDSS, VIPERS, and VVDS, and
when the demographics of sources in a low-quality training set
do not match the demographics of objects in the test data set.
We demonstrate that our Bayesian combination technique improves
the overall performance over any individual classification method
in these scenarios.
Thus, strategies that combine the predictions of different classifiers
may prove to be optimal in currently ongoing and forthcoming
photometric surveys,
such as the Dark Energy Survey and the Large Synoptic Survey Telescope.

@@

The problem of source classification is fundamental to astronomy
and goes as far back as \citet{messier1781catalogue}.
A variety of different strategies have been developed 
to tackle this long-standing problem,
and yet there is no consensus on
the optimal star-galaxy classification strategy.
The most commonly used method to classify stars and galaxies
in large sky surveys is the morphological separation
~\citep{sebok1979optimal, kron1980photometry, valdes1982resolution,
yee1991faint, vasconcellos2011decision,
henrion2011bayesian}.
It relies on the assumption that
stars appear as point sources
while galaxies appear as resolved sources.
However,
currently ongoing and upcoming large photometric surveys,
such as the Dark Energy Survey
(DES\footnote{http://www.darkenergysurvey.org/})
and the Large Synoptic Survey Telescope
(LSST\footnote{http://www.lsst.org/lsst/}),
will detect a vast number of unresolved galaxies
at faint magnitudes.
Near a survey's limit, the photometric observations
cannot reliably separate stars from unresolved galaxies
by morphology alone without leading to
incompleteness and contamination in the star and galaxy samples.

@@

A self-organizing map~\citep{kohonen1990self, kohonen2001self}
is an unsupervised, artificial neural network algorithm
that is capable of projecting high-dimensional input data
onto a low-dimensional map
through a process of competitive learning.
In astronomical applications,
the high-dimensional input data can be
magnitudes, colors, or some other photometric attributes.
The output map is usually chosen to be two-dimensional
so that the resulting map can be used for visualizing
various properties of the input data.
The differences between a SOM and
typical neural network algorithms are
that a SOM is unsupervised,
there are no hidden layers and therefore no extra parameters,
and it produces a direct mapping
between the training set and the output network.
In fact, a SOM can be viewed as a non-linear generalization
of a principal component analysis (PCA) algorithm~\citep{yin2008self}.

@@

During the construction of a SOM,
each node on the two-dimensional map is represented by
weight vectors of the same dimension
as the number of attributes used to create the map itself.
In an iterative process,
each object in the input sample is individually used
to correct these weight vectors.
This correction is determined so that the specific neuron (or node),
which at a given moment best represents the input source,
is modified along with the weight vectors
of that node's neighboring neurons.
As a result, this sector within the map
becomes a better representation of the current input object.
This process is repeated for every object in the training data,
and the entire process is repeated for several iterations.
Eventually, the SOM converges to its final form where
the training data is separated into groups of similar features.
Although the spectroscopic labels are not used at all
in the learning process, they are used
(only after the map has been constructed)
to generate predictions for each cell in the resulting two-dimensional map.

@@

One of the most important parameter in SOMc
is the topology of the two-dimensional SOM,
which can be rectangular, hexagonal, or spherical.
In our SOM implementation, it is also possible
to use periodic boundary conditions for the non-spherical cases.
The spherical topology is by definition periodic and
is constructed by using
\textsc{HEALPIX}~\citep{gorski2005healpix}.
Similar to TPC,
we use the OOB technique to make an unbiased estimation of errors.
We determine the optimal parameters by performing
a grid search in the parameter space of different toplogies,
as well as other SOM parameters, for the OOB data.
We find that the spherical topology gives the best performance
for the CFHTLenS data, likely due to its natural periodicity.
Thus, we use a spherical topology to classify stars and galaxies
in the CFHTLenS data.
For a complete description of the SOM implementation
and its application to the estimation of
photo-$z$ probability density functions (photo-$z$ PDFs),
we refer the reader to \cite{carrascokind2014somz}.

@@

We therefore adopt a binning strategy similar to 
\cite{carrascokind2014exhausting}.
In this binning strategy,
we allow different classifier combinations 
in different parts of parameter space
by creating two-dimensional SOM representations of
the full nine-dimensional magnitude-color space:
$u$, $g$, $r$, $i$, $z$, $u-g$, $g-r$, $r-i$, and $i-z$.
A SOM representation can be rectangular, hexagonal, or spherical;
here we choose a 10$\times$10 rectangular topology to facilitate 
visualization as shown in Figure~\ref{fig:som_colors}.
We note that this choice is mainly for convenience
and that the optimal topology and map size would likely depend on
a number of factors, such as the number of objects and attributes.
For all combination methods,
we use only the OOB (cross-validation) data contained in each cell
to compute the relative weights for the base classifiers.
The weights within individual cells are then applied to
the blind test data set to make the prediction.

@@

We have cross-matched reliable spectroscopic galaxies from
the Deep Extragalactic Evolutionary Probe Phase 2~
\citep[DEEP2;][]{davis2003science,newman2013deep2},
the Sloan Digital Sky Survey Data Release 10~\citep[SDSS-DR10]{Ahn2014},
the VIsible imaging Multi-Object Spectrograph (VIMOS)
Very Large Telescope (VLT) Deep Survey~
\citep[VVDS;][]{le2005vimos,garilli2008vimos}, and
the VIMOS Public Extragalactic Redshift
Survey~\citep[VIPERS;][]{garilli2014vimos}.
We have selected only sources with very secure
redshifts and no bad flags (quality flags -1, 3, and 4 for DEEP2;
quality flag 0 for SDSS; quality flags 3, 4, 23, and 24 for VIPERS
and VVDS).
In the end, we have 8,545 stars and 57,843 galaxies available
for the training and testing processes.
We randomly select 13,278 objects for the blind testing set,
and use the remainder for training and cross-validation.
While HB uses only the magnitudes in the five bands,
$u$, $g$, $r$, $i$, and $z$,
TPC and SOMc are trained with a total of 9 attributes: the five magnitudes
and their corresponding colors, $u-g$, $g-r$, $r-i$, and $i-z$.
The morphological separation method uses \textsc{SExtractor}'s 
FLUX\_RADIUS parameter provided by the CFHTLenS catalog.

Our goal here is not to obtain the best classifier
performance; for this we would have fine tuned individual base
classifiers and chosen sophisticated models best suited to the
particular properties of the CFHTLenS data.
For example, \cite{hildebrandt2012cfhtlens} suggest that
all objects with $i > 23$ in the CFHTLenS data set
may be classified as galaxies
without significant incompleteness and contamination in the galaxy sample.
Although this approach works because
the high Galactic latitude fields of the CFHTLS contain
relatively few stars,
it is very unlikely that such an approach
will meet the science requirements for the quality of
star-galaxy classification in lower-latitude, star-crowded fields.
Rather, our goal for the CFHTLenS data set
is to demonstrate the usefulness
of combining different classifiers even when
the base classifiers may be poor or trained on partial data.
We also note that the relatively few number of stars
in the CFHTLS fields might paint too positive a picture of completeness and purity,
especially for the stars.
Thus, we caution the reader that the specific completeness and purity values
will likely vary in other surveys that observe large portions of the sky,
and we emphasize once again that our aim is to highlight that there is a
relative improvement in performance when we combine multiple star-galaxy
classification techniques to generate a meta-classification.

@@

%% rich training data
\begin{table*}
  \caption{A summary of the classification performance metrics
           for the four individual methods
           and the four different classification combination methods
           as applied to the CFHTLenS data,
           with no cut applied to the training data set.
           The definition of the metrics is summarized in
           Table~\ref{table:metrics}.
           The bold entries highlight the best performance values
           within each column.
           Note that some objects in the test set have bad or missing
           values (\eg $-99$ or $99$) in one or more attributes,
           which are included here (but are omitted, for example,
           in Figure~\ref{fig:purity_mag_integrated} when the corrsponding
           attribute is not available.)}
  \centering
  \input{tables/metrics_all.tex}
  \label{table:metrics_all}
\end{table*}

@@

In Figure~\ref{fig:purity_mag}, 
we compare the star and galaxy purity values
for BMC, TPC, and morphological separation
as functions of $i$-band magnitude.
We use the kernel density estimation \cite[KDE;][]{silverman1986density}
with the Gaussian kernel to smooth the fluctuations in the distribution.
Although morphological separation shows a slightly better performance
in galaxy purity at bright magnitudes $i \la 20$,
BMC outperforms both TPC and morphological separation
at faint magnitudes $i \ga 21$.
As the top panel shows,
the number count distribution peaks at $i \sim 22$,
and BMC therefore outperforms both TPC and morphological separation
for the majority of objects.
It is also clear that BMC outperforms TPC over all magnitudes.
BMC can presumably accomplish this by combining information from
all base classifiers,
\eg giving more weight to the morphological separation method
at bright magnitudes.
The bottom panel shows that
the star purity of morphological separation drops to $p_s < 0.8$
at fainter magnitudes $i > 21$.
This is expected, as our crude morphological separation classifies
every object as a galaxy beyond $i > 21$, and
purity measures the number of true stars classified as stars.
It is again clear that BMC outperforms both TPC and morphological separation
in star purity values over all magnitudes.

In Figure~\ref{fig:purity_mag_integrated}, we show the
cumulative galaxy and star purity values as functions of magnitude.
Although morphological separation performs better than TPC
at bright magnitudes, its purity values decrease as 
the magnitudes become fainter, and
TPC eventually outperforms morphological separation by 1--2\% at
$i > 21$.
BMC clearly outperforms both TPC and morphological separation,
and it maintains the overall galaxy purity of 0.980
up to $i \sim 24.5$.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/purity_mag.pdf}
  \caption{Purity as a function of the $i$-band magnitude
           as estimated by the kernel density estimation (KDE) method.
           The top panel shows the histogram with a bin size of 0.1 mag
           and the KDE for objects in the test set.
           The second panel shows the fraction of stars estimated by KDE
           as a function of magnitude.
           The bottom two panels compare
           the galaxy and star purity values for BMC, TPC, and 
           morphological separation as functions of magnitude.
           Results for BMC, TPC, and morphological separation are in
           blue, green, and red, respectively.
           The $1 \sigma$ confidence bands are estimated by
           bootstrap sampling.}
  \label{fig:purity_mag}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/purity_mag_integrated.pdf}
  \caption{Cumulative purity as a function of the $i$-band magnitude.
           The upper panel compares
           the galaxy purity values for BMC (blue solid line),
           TPC (green dashed line), and 
           morphological separation (red dashed line).
           The lower panel compares the star purity.
           The $1 \sigma$ error bars are computed following the method
           of \citet{paterno2004calculating} to avoid the unphysical
           errors of binomial or Poisson statistics.}
  \label{fig:purity_mag_integrated}
\end{figure}

@@

\begin{table*}
  \caption{A summary of the classification performance metrics
           for the four individual methods
           and the four different classification combination methods
           when the training data set consists of
           only the sources that are in CFHTLS W1 field,
           has spectroscopic labels available from VVDS,
           and has $i < 22$.
           The definition of the metrics is summarized in
           Table~\ref{table:metrics}.
           The bold entries highlight the best performance values
           within each column.
           Note that some objects in the test set have bad or missing
           values (\eg $-99$ or $99$) in one or more attributes,
           which are included here (but are omitted, for example,
           in Figure~\ref{fig:purity_mag_cut_integrated} when the corrsponding
           attribute is not available.)}
  \centering
  \input{tables/metrics_cut.tex}
  \label{table:metrics_cut}
\end{table*}

We present in Table~\ref{table:metrics_cut}
the same six metrics for each method,
and highlight the best method for each metric.
Overall, the results obtained for the reduced data set are remarkable.
With a smaller training set, our training based methods, TPC and SOMc,
suffer a significant decrease in performance.
The performance of morphological separation and HB
is essentially unchanged from Table~\ref{table:metrics_all}
as they do not depend on the training data.
Without sufficient training data,
the advantage of combining the predictions of different classifiers
is more obvious.
Even WA, the simplest of combination techniques, outperforms
all individual classification techniques in four metrics,
AUC, $p_s$ at $c_s=0.7145$, $p_g$ at $c_g=0.9600$, and $p_s$ at $c_s=0.2500$.
Although BoM always chooses TPC as the best model
when we have a high-quality training set,
it now chooses various methods in different bins
and outperforms all base classifiers.
While the performance of the stacking technique is only slightly worse
than that of BMC when we have a high-quality training set,
stacking now fails to outperform morphological separation.
BMC shows an impressive performance and 
outperforms all other classification techniques in all six metrics.
Overall, the improvements are small but still significant
since these metrics are averaged over the full test data.

@@

In Figure~\ref{fig:weights_cut}, we again show the $10\times10$
two-dimensional weight map defined by the SOM.
When the quality of training data is relatively poor,
the performance of training based algorithms will decrease,
while the performance of template fitting algorithms
or morphological separation methods
is independent of training data.
Thus, when the weight maps of Figure~\ref{fig:weights}
and Figure~\ref{fig:weights_cut} are visually compared,
it is clear that
the BMC algorithm now uses more information from
morphological separation and HB,
while it uses considerably less information from 
our training based algorithms, TPC and SOMc.
Not surprisingly, the morphological separation method
performs best at bright magnitudes,
and BMC assigns more weight to HB at fainter magnitudes.

We present the star and galaxy purity values as functions of
$i$-band magnitude in Figure~\ref{fig:purity_mag_cut}.
The normalized density distribution as a function of magnitude
in the top panel
and the stellar distribution in the second panel
clearly show that the demographics of the training set and 
that of the test set are different.
Since the training set is cut at $i < 22$,
the density distribution falls off sharply around $i \sim 22$
and has a higher fraction of stars than the test set.
Compared to the purity values in Figure~\ref{fig:purity_mag},
TPC now suffers a significant decrease in star and galaxy purity.
However, the purity of BMC does not show such a significant drop
and decreases by only 2--5\%.
As suggested by the weight maps in Figure~\ref{fig:weights_cut},
BMC can accomplish this by shifting the relative weights assigned to
each base classifier in different SOM cells.
As the quality of training set worsens,
BMC assigns less weight to training based methods
and more weight to HB and morphological separation.

In Figure~\ref{fig:purity_mag_cut_integrated}, we show the
cumulative galaxy and star purity values as functions of magnitude.
Compared to Figure~\ref{fig:purity_mag_integrated},
the drop in the performance of TPC is clear.
However, even when some classifiers have been trained on
a significantly reduced training set,
BMC maintains a galaxy purity of 0.970 and a star purity of 1.0
up to $i \sim 24.5$, and it sill outperforms morphological separation
at fainter magnitudes $i \ga 21$.

@@

@@

Similarly, we impose various spectroscopic redshift cuts 
on the training data in Figure~\ref{fig:perform_z_cut}.
Since all stars have $z_{\rmn{spec}}$ values
close to zero, we are effectively changing the demographics
of training set by keeping all stars and gradually removing
galaxies with high redshifts.
BMC begins to perform worse than morphological separation
when a conservative cut of $z_{\rmn{spec}} < 0.6$ is imposed.
However, it is again clear that BMC is able to
utilize information from HB and morphological separation to
mitigate the drop in the performance of TPC.

@@

The problem of star-galaxy classification is a rich area
for future research.
It is unclear if sufficient training data will be available in future
ground-based surveys. Furthermore, in large sky surveys such as DES and LSST,
photometric quality is not uniform across the sky,
and a purely morphological classifier alone will not be sufficient,
especially at faint magnitudes.
Given the efficacy of our approach,
classifier combination strategies are likely the optimal approach
for currently ongoing and forthcoming photometric surveys.
We therefore plan to apply the combination technique described in this paper
to other surveys such as the DES.
Our approach can also be extended more broadly to
classify objects that are neither stars nor galaxies (\eg quasars).
Finally, future studies could explore the use of multi-epoch data,
which would be particularly useful for the next generation of
synoptic surveys.

@@

The authors thank the referee for a careful reading of the manuscript
and comments that improved this work.
We thank Ignacio Sevilla for helpful and insightful conversations.
We acknowledge support from the 
National Science Foundation Grant No.\ AST-1313415.
RJB acknowledges support as an Associate
within the Center for Advanced Study at the University of Illinois.
